### 2024-04-05 sections/2-methods.tex
compiles, switched to algorithmicx, tables still bad
paper edits up until section 4.3

In this manuscript, we are primarily concerned with $k$-NN search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$ of cardinality $|\textbf{X}| = n$, we define a \textit{point} or \textit{datum} $x_i \in \textbf{X}$ as a singular observation, for example, the neural-network embedding of an image, the genome of an organism, a measurement of a radio frequency signal, etc.

@@

The choice of distance function varies by dataset and domain.
For example, with neural-network embeddings, one could use Euclidean (L2) or Cosine distance.
With genomic or proteomic data, Levenshtein, Smith-Waterman, Needleman-Wunsch or Hamming distances are useful.
With time-series data, one could use Dynamic Time Warping (DTW) or Wasserstein distance.

CAKES assumes the manifold hypothesis~\cite{fefferman2016testing}, i.e., high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space.
We say that such data are \textit{manifold-constrained} and have low \textit{local fractal dimension} (LFD).
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$.
While we sometimes use Euclidean notions, such as voids, volumes and proximity to describe the geometric and topological properties of the clusters and manifold, CLAM and CAKES do not rely on such notions;
they serve merely as convenient and intuitive vocabulary to discuss the underlying mathematics.

@@

\begin{equation}
    \frac{\text{log} \left( \frac{|B_X(q, r_1)|}{|B_X(q, r_2)|} \right) }{\text{log} \left( \frac{r_1}{r_2} \right) }
    \label{eq:methods:lfd-original}
\end{equation}
where $B_X(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases.

@@

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The \textit{center} is the geometric median of the points (or a smaller sample of the points) in the \textit{cluster}, and so it is a real data point.
The \textit{radius} is the maximum distance from the \textit{center} to any point in the \textit{cluster}.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a subset of the metric ball at the same center and radius, i.e. $C(c, r) \subseteq B_X(c, r)$.

@@

The \textit{metric entropy} $\mathcal{N}_{r}(X)$ for some radius $r$ was defined, in~\cite{yu2015entropy}, as the minimum number of clusters of a uniform radius $r$ needed to cover the data.
In this paper, we define the metric entropy $\mathcal{N}_{\hat{r}}(X)$ of a dataset $X$ for the hierarchical clustering scheme as the number of leaf clusters in the tree where $\hat{r}$ is the mean radius of all leaf clusters.

@@

The \textit{radius} of $C$ is the maximum distance from the \textit{center} to any other point in $C$.
The point which is responsible for that radius (i.e., the furthest point from the center) is designated the \textit{left pole} and the point which is furthest from left pole is designated the \textit{right pole}.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Partition($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $C$, a cluster
        \REQUIRE $criteria$, user-specified stopping criteria

        \STATE $seeds \Leftarrow$ random sample of $\left\lceil \sqrt{|C|} \right\rceil$ points from $C$
        \STATE $c \Leftarrow$ geometric median of $seeds$
        \STATE $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \STATE $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \STATE $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \STATE $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$

        \IF{$|L| > 1$ \textbf{and} $L$ satisfies $criteria$}
            \STATE Partition($L$, $criteria$)
        \ENDIF

        \IF{$|R| > 1$ \textbf{and} $R$ satisfies $criteria$}
            \STATE Partition($R$, $criteria$)
        \ENDIF
    \end{algorithmic}
\end{algorithm}

@@

Given Algorithm~\ref{alg:methods:partition}, we can now pose the $k$-NN and $\rho$-NN search problems.

@@

We define a \textit{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains duplicates of the same point (i.e., has cardinality greater than 1 but contains only one \textit{unique} point).
A singleton clearly has zero radius, and so $\delta = \delta^{-} = \delta^{+}$.
Hence, we overload the above notation to also refer to the distance from a query to an individual point.

@@

We conduct $\rho$-NN search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of always proceeding with both of its children, we proceed only with those children which might contain points in the query ball.

@@

To determine whether both children can contain points in the query ball, we consider Figure~\ref{fig:methods:overlapping-children}.
Here, we overload the notation for $\overline{x y}$ to refer to the line segment joining points $x$ and $y$ as well as the length of that line segment.

Let $q$ denote the query, $\rho$ denote the search radius, and $l$ and $r$ denote the cluster's left and right poles respectively (see Section~\ref{sec:methods:clustering:building-the-tree}).
Without loss of generality, we assume that $\overline{q r \vphantom{l}} \leq \overline{q l}$.
Now let $q'$ be the projection of $q$ to $\overline{l r}$, $m$ be the midpoint of $\overline{l r}$, and $d$ be the distance from $q'$ to $m$.
As a consequence of how we assign a point in the parent cluster to the left child in Algorithm~\ref{alg:methods:partition}, if $q'$ is too far from $m$, i.e. $\rho < d$, then when the left child cannot contain points inside the query ball.
In such a case we proceed only with the right child.
Otherwise, we proceed with both children.

@@

\begin{algorithm} 
    \caption{tree-search($C$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search:tree-search} 
    \begin{algorithmic}
        \REQUIRE $C$, a cluster
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

        \IF{$C$ is a leaf \textbf{or} $\delta^+_C \leq \rho$}
            \STATE \textbf{return} $\{C\}$
        \ELSE

            \STATE $[l, r]$ $\Leftarrow$ \textit{pole}s of $C$
            \STATE $[L, R]$ $\Leftarrow$ \textit{children} of $C$

            \IF{$\overline{q l} < \overline{q r \vphantom{l}}$}
                \STATE $[r, l] \Leftarrow [l, r]$
                \STATE $[R, L] \Leftarrow [L, R]$
            \ENDIF

            \IF{$(\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho$}
                \STATE \textbf{return} tree-search($L, q, \rho$) $\cup$ tree-search($R, q, \rho$)
            \ELSE
                \STATE \textbf{return} tree-search($R, q, \rho$)
            \ENDIF
        \ENDIF
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{leaf-search($S$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search:leaf-search} 
    \begin{algorithmic}
        \REQUIRE $S$, a set of clusters
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

        \STATE $H \Leftarrow \emptyset$, a set of hits

        \FOR{$C \in S$}
            \IF{$\delta^+_C \leq \rho$}
                \STATE $H$ $\Leftarrow$ $H \cup \{C\}$
            \ELSE
                \FOR{$p \in C$}
                    \IF{$f(p, q) \leq \rho$}
                        \STATE $H$ $\Leftarrow$ $H \cup \{p\}$
                    \ENDIF
                \ENDFOR
            \ENDIF
        \ENDFOR

        \STATE \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{$\rho$-NN-search($root$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $\rho$, a search radius

        \STATE $S$ $\Leftarrow$ tree-search($root$, $q$, $\rho$)
        \STATE $H$ $\Leftarrow$ leaf-search($S$, $q$, $\rho$)
        \STATE \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $k$, the number of neighbors to find

        \STATE $r \Leftarrow radius$ of the $root$ cluster
        \STATE $r \Leftarrow$ $\frac{r}{|root|}$

        \ALOOP{}
            \STATE $Q \Leftarrow$ tree-search($root$, $q$, $r$)
            \IF{$Q \neq \emptyset$}
                \STATE \textbf{break}
            \ENDIF
            \STATE $r \Leftarrow 2 \cdot r$
        \ENDALOOP

        \ALOOP{}
            \IF{$\sum_{C \in Q} |C| >= k$}
                \STATE \textbf{break}
            \ENDIF
            \STATE $\mu \Leftarrow \frac{|S|}{\sum_{C \in Q} \frac{1}{LFD(C)}}$
            \STATE $r \Leftarrow r \cdot \min \left( 2, \left( {\frac{k}{\sum_{C \in Q} |C|}} \right)^{\mu^{-1}} \right)$
            \STATE $S \Leftarrow$ tree-search($root$, $q$, $r$)
        \ENDLOOP

        \STATE $H \Leftarrow$ sort($\rho$-NN-search($root$, $q$, $r$))
        \STATE \textbf{return} $H[.. k]$
    \end{algorithmic}
\end{algorithm}

@@

where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset, $d$ is the LFD of the dataset, and $k$ is the number of nearest neighbors.
We note that the scaling factor should be close to 1 unless fractal dimension is highly variable in a small region (i.e. if $\hat{d}$ differs significantly from $d$).

@@

We then use the QuickSelect algorithm, modified to account for multiplicities and to reorder $Q$ in-place, to find the element in $Q$ with the $k^{th}$ smallest $\delta^{+}$; in other words, we find $\tau$, the smallest $\delta^{+}$ in $Q$ such that $\left| B_X(q, \tau) \right| \geq k$.
Since this step may require a binary search for the correct pivot element to find $\tau$ and reordering with a new pivot is linear-time, this version of QuickSelect may require $\mathcal{O}(|Q| \log |Q|)$ time.

@@

We continue this process until $Q$ no longer contains any clusters.
We then use the QuickSelect algorithm once last time to reorder $Q$, find $\tau$, and return the $k$ nearest neighbors.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Breadth-First Sieve($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:bredth-first-sieve} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $k$, the number of neighbors to find

        \STATE $c \Leftarrow$ \textit{center} of $root$
        \STATE $Q \Leftarrow$ \{ ($root$, $\delta^{+}_{root}$, $|root| - 1$), ($c$, $\delta_{root}$, 1) \}

        \ALOOP{}
            \IF{$k = \sum_{(\_, \_, m) \in Q} m$}
                \STATE \textbf{break}
            \ENDIF
            \STATE $\tau \Leftarrow$ QuickSelect($Q$, $k$)
            \FOR{$(p, \delta^{+}_{p}, m) \in Q$}
                \IF{$\delta^{-}_{p} > \tau$}
                    \STATE $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                \ENDIF
            \ENDFOR
            \FOR{$(p, \delta^{+}_{p}, m) \in Q$}
                \STATE $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                \IF{$p$ is a point}
                    \STATE \textbf{continue}
                \ELSIF{$p$ is a leaf}
                    \FOR{$c \in p$}
                        \STATE $Q \Leftarrow Q \cup \{ (c, \delta_{c}, 1) \}$
                    \ENDFOR
                \ELSE
                    \STATE $[L, R] \Leftarrow$ children of $p$
                    \STATE $Q \Leftarrow Q \cup \{ (L, \delta^{+}_{L}, |L| - 1), (L_c, \delta_{L}, 1) \}$
                    \STATE $Q \Leftarrow Q \cup \{ (R, \delta^{+}_{R}, |R| - 1), (R_c, \delta_{R}, 1) \}$
                \ENDIF
            \ENDFOR
        \ENDALOOP

        \STATE \textbf{return} $Q$
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm} 
    \caption{Depth-First Sieve($root$, $q$, $k$)} 
    \label{alg:methods:depth-first-sieve} 
    \begin{algorithmic}
        \REQUIRE $root$, the root cluster
        \REQUIRE $q$, a query
        \REQUIRE $k$, the number of neighbors to find

        \STATE{$Q \Leftarrow [root]$, a min-priority queue by $\delta^{-}$}
        \STATE{$H \Leftarrow []$, a max-priority queue by $\delta$}

        \WHILE{$|H| < k$ \textbf{or} $H.peek.\delta \geq Q.peek.\delta^{-}$}
            \WHILE{$\neg (Q.peek$ is a leaf)}
                \STATE{$C \Leftarrow Q.pop$, the closest cluster}
                \STATE{$[L, R] \Leftarrow$ children of $C$}
                \STATE{$Q.push(L)$}
                \STATE{$Q.push(R)$}
            \ENDWHILE
            \STATE{$leaf \Leftarrow Q.pop$}
            \FOR{$p \in leaf$}
                \STATE{$H.push(p)$}
            \ENDFOR
            \WHILE{$|H| > k$}
                \STATE{$H.pop$}
            \ENDWHILE
        \ENDWHILE
        \STATE Return $H$
    \end{algorithmic}
\end{algorithm}

@@

We combine the complexity analyses of the two Sieve methods because they are very similar.
For these methods we, once again, use the terminology of tree-search and leaf-search.
Tree-search navigates the cluster tree and adds clusters to $Q$.
Leaf-search exhaustively searches some of the clusters in $Q$ to find the $k$ nearest neighbors.

@@

We perform some na\"{i}ve auto-tuning to select the optimal $k$-NN algorithm to use with a given dataset.
We start by taking the center of every cluster at a low depth, e.g. 10, in the cluster tree as a query.
This gives us a small, representative sample of the dataset.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the three algorithms described in Section~\ref{sec:methods:knn-search}.
We select the fastest algorithm over all the queries as the optimal algorithm for that dataset and value of $k$.
Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.

@@

To elaborate on the augmentation process, we start with an original dataset from the ANN-benchmarks suite.
Let $X$ be the dataset, $d$ be its dimensionality, $\epsilon$ be a user-specified noise level and $m$ be a user-specified integer multiplier.
For each datum $x \in X$, we create $m - 1$ new data points within a distance $\epsilon \cdot ||x||$ of $x$ where $||x||$ is the Euclidean distance from $x$ to the origin.
We construct a random vector $r$ of $d$ dimensions in the hyper-sphere of radius $\epsilon$ centered at the origin.
We then add $r$ to $x$ to get a new point $x'$.
Since $||r|| \leq \epsilon$, we have that $||x - x'|| \leq \epsilon$ (i.e., $x'$ is within a distance $\epsilon$ of $x$).
This produces a new dataset $X'$ with $|X'| = m \cdot |X|$.

>>>>

In this manuscript, we are primarily concerned with $k$-NN search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$ of cardinality $|\textbf{X}| = n$, we define a \textit{point} or \textit{datum} $x_i \in \textbf{X}$ as a singular observation. Examples include the neural-network embedding of an image, the genome of an organism, and a measurement of a radio frequency signal.

@@

The choice of an appropriate distance function varies by dataset and domain.
For example, with neural-network embeddings, one could use Euclidean (L2) or Cosine distance.
With genomic or proteomic data, Levenshtein, Smith-Waterman, Needleman-Wunsch or Hamming distances are useful.
With time-series data, one could use Dynamic Time Warping (DTW) or Wasserstein distance.

CAKES assumes the manifold hypothesis~\cite{fefferman2016testing}, the notion that high-dimensional data collected from constrained generating phenomena typically only occupy a low-dimensional manifold within their embedding space.
We say that such data are \textit{manifold-constrained} and have low \textit{local fractal dimension} (LFD).
In other words, we assume that the dataset is embedded in a $D$-dimensional space, but that the data only occupy a $d$-dimensional manifold, where $d \ll D$.
While we sometimes use Euclidean notions, such as voids, volumes and proximity to describe the geometric and topological properties of the clusters and manifold, CLAM and CAKES do not rely on such notions;
they serve merely as convenient and intuitive vocabulary to discuss the underlying mathematics.

@@

\begin{equation}
    \frac{\text{log} \left( \frac{|B_X(q, r_1)|}{|B_X(q, r_2)|} \right) }{\text{log} \left( \frac{r_1}{r_2} \right) }
    \label{eq:methods:lfd-original}
\end{equation}
where $B_X(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can overload terminology to say that the dataset has low LFD.

@@

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The \textit{center} is the geometric median of the points (or a smaller sample of the points) in the \textit{cluster}, and so it is a real data point.
The \textit{radius} is the maximum distance from the \textit{center} to any point in the \textit{cluster}.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a proper subset of the metric ball at the same center and radius, i.e. $C(c, r) \subset B_X(c, r)$.

@@

The \textit{metric entropy} $\mathcal{N}_{r}(X)$ for some radius $r$ was defined, in~\cite{yu2015entropy}, as the minimum number of clusters of a uniform radius $r$ needed to cover the data.
In this paper, where the clustering is hierarchical rather than flat, we define the metric entropy $\mathcal{N}_{\hat{r}}(X)$ of a dataset $X$ as the number of leaf clusters in the tree where $\hat{r}$ is the mean radius of all leaf clusters.

@@

The \textit{radius} of $C$ is the maximum distance from the \textit{center} to any other point in $C$.
The point which is responsible for that radius (i.e., the furthest point from the center) is designated the \textit{left pole} and the point which is furthest from the left pole is designated the \textit{right pole}.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Partition($C$, $criteria$)} % give the algorithm a caption
    \label{alg:methods:partition} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[0] % enter the algorithmic environment
        \Require $C$, a cluster
        \Require $criteria$, user-specified stopping criteria

        \State $seeds \Leftarrow$ random sample of $\left\lceil \sqrt{|C|} \right\rceil$ points from $C$
        \State $c \Leftarrow$ geometric median of $seeds$
        \State $l \Leftarrow \argmax f(c, x) \ \forall \ x \in C$
        \State $r \Leftarrow \argmax f(l, x) \ \forall \ x \in C$
        \State $L \Leftarrow \{x \ | \ x \in C \land f(l, x) \le f(r, x)\}$
        \State $R \Leftarrow \{x \ | \ x \in C \land f(r, x) < f(l, x)\}$

        \If{$|L| > 1$ \textbf{and} $L$ satisfies $criteria$}
            \State Partition($L$, $criteria$)
        \EndIf

        \If{$|R| > 1$ \textbf{and} $R$ satisfies $criteria$}
            \State Partition($R$, $criteria$)
        \EndIf
    \end{algorithmic}
\end{algorithm}

@@

Given Algorithm~\ref{alg:methods:partition}, we can now formally pose the $k$-NN and $\rho$-NN search problems.

@@

We define a \textit{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains only duplicates of the same point (i.e., has cardinality greater than 1 but contains only one \textit{unique} point).
A singleton clearly has zero radius, and so $\delta = \delta^{-} = \delta^{+}$.
Hence, we overload the above notation to also refer to the distance from a query to an individual point.

@@

We conduct $\rho$-NN search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of always proceeding to search both of its children, we proceed only with those children which might contain points in the query ball.

@@

To determine whether both children can contain points in the query ball, we consider Figure~\ref{fig:methods:overlapping-children}.
Here, we overload the notation for $\overline{x y}$ to refer both to the line segment joining points $x$ and $y$ as well as to the length of that line segment.

Let $q$ denote the query, $\rho$ denote the search radius, and $l$ and $r$ denote the cluster's left and right poles respectively (see Section~\ref{sec:methods:clustering:building-the-tree}).
Without loss of generality, we assume that $\overline{q r \vphantom{l}} \leq \overline{q l}$.
Now let $q'$ be the projection of $q$ onto $\overline{l r}$, $m$ be the midpoint of $\overline{l r}$, and $d$ be the distance from $q'$ to $m$.
As a consequence of how we assign a point in the parent cluster to the left child in Algorithm~\ref{alg:methods:partition}, if $\rho < d$, then the left child cannot contain points inside the query ball.
In such a case we proceed to search only the right child.
Otherwise, we proceed with both children.

@@

\begin{algorithm} 
    \caption{tree-search($C$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search:tree-search} 
    \begin{algorithmic}[0]
        \Require $C$, a cluster
        \Require $q$, a query
        \Require $\rho$, a search radius

        \If{$C$ is a leaf \textbf{or} $\delta^+_C \leq \rho$}
            \State \textbf{return} $\{C\}$
        \Else

            \State $[l, r]$ $\Leftarrow$ \textit{pole}s of $C$
            \State $[L, R]$ $\Leftarrow$ \textit{children} of $C$

            \If{$\overline{q l} < \overline{q r \vphantom{l}}$}
                \State $[r, l] \Leftarrow [l, r]$
                \State $[R, L] \Leftarrow [L, R]$
            \EndIf

            \If{$(\overline{q l} + \overline{q r \vphantom{l}})(\overline{q l} - \overline{q r \vphantom{l}}) \leq 2 \cdot \overline{l r} \cdot \rho$}
                \State \textbf{return} tree-search($L, q, \rho$) $\cup$ tree-search($R, q, \rho$)
            \Else
                \State \textbf{return} tree-search($R, q, \rho$)
            \EndIf
        \EndIf
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{leaf-search($S$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search:leaf-search} 
    \begin{algorithmic}[0]
        \Require $S$, a set of clusters
        \Require $q$, a query
        \Require $\rho$, a search radius

        \State $H \Leftarrow \emptyset$, a set of hits

        \For{$C \in S$}
            \If{$\delta^+_C \leq \rho$}
                \State $H$ $\Leftarrow$ $H \cup \{C\}$
            \Else
                \For{$p \in C$}
                    \If{$f(p, q) \leq \rho$}
                        \State $H$ $\Leftarrow$ $H \cup \{p\}$
                    \EndIf
                \EndFor
            \EndIf
        \EndFor

        \State \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm} 
    \caption{$\rho$-NN-search($root$, $q$, $\rho$)} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $\rho$, a search radius

        \State $S$ $\Leftarrow$ tree-search($root$, $q$, $\rho$)
        \State $H$ $\Leftarrow$ leaf-search($S$, $q$, $\rho$)
        \State \textbf{return} $H$
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[0] % enter the algorithmic environment
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

        \State $r \Leftarrow radius$ of the $root$ cluster
        \State $r \Leftarrow$ $\frac{r}{|root|}$

        \Loop
            \State $Q \Leftarrow$ tree-search($root$, $q$, $r$)
            \If{$Q \neq \emptyset$}
                \State \textbf{break}
            \EndIf
            \State $r \Leftarrow 2 \cdot r$
        \EndLoop

        \Loop
            \If{$\sum_{C \in Q} |C| >= k$}
                \State \textbf{break}
            \EndIf
            \State $\mu \Leftarrow \frac{|S|}{\sum_{C \in Q} \frac{1}{LFD(C)}}$
            \State $r \Leftarrow r \cdot \min \left( 2, \left( {\frac{k}{\sum_{C \in Q} |C|}} \right)^{\mu^{-1}} \right)$
            \State $S \Leftarrow$ tree-search($root$, $q$, $r$)
        \EndLoop

        \State $H \Leftarrow$ sort($\rho$-NN-search($root$, $q$, $r$))
        \State \textbf{return} $H[.. k]$
    \end{algorithmic}
\end{algorithm}

@@

where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset, $d$ is the LFD of the dataset, and $k$ is the number of nearest neighbors.
We note that the scaling factor should be close to 1 unless fractal dimension is highly variable in the region around the query (i.e. if $\hat{d}$ differs significantly from $d$).

@@

We then use the QuickSelect algorithm, modified to account for multiplicities and to reorder $Q$ in-place, to find the element in $Q$ with the $k^{th}$ smallest $\delta^{+}$; in other words, we find $\tau$, the smallest $\delta^{+}$ in $Q$ such that $\left| B_X(q, \tau) \right| \geq k$.
Since this step may require a binary search for the correct pivot element to find $\tau$ and reordering with a new pivot takes linear time in the size of the input list, this version of QuickSelect has $\mathcal{O}(|Q| \log |Q|)$ time complexity.

@@

We continue this process until $Q$ no longer contains any clusters.
We then use the QuickSelect algorithm one last time to reorder $Q$, find $\tau$, and return the $k$ nearest neighbors.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Breadth-First Sieve($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:bredth-first-sieve} % and a label for \ref{} commands later in the document
    \begin{algorithmic}[0] % enter the algorithmic environment
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

        \State $c \Leftarrow$ \textit{center} of $root$
        \State $Q \Leftarrow$ \{ ($root$, $\delta^{+}_{root}$, $|root| - 1$), ($c$, $\delta_{root}$, 1) \}

        \Loop
            \If{$k = \sum_{(\_, \_, m) \in Q} m$}
                \State \textbf{break}
            \EndIf
            \State $\tau \Leftarrow$ QuickSelect($Q$, $k$)
            \For{$(p, \delta^{+}_{p}, m) \in Q$}
                \If{$\delta^{-}_{p} > \tau$}
                    \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                \EndIf
            \EndFor
            \For{$(p, \delta^{+}_{p}, m) \in Q$}
                \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                \If{$p$ is a point}
                    \State \textbf{continue}
                \ElsIf{$p$ is a leaf}
                    \For{$c \in p$}
                        \State $Q \Leftarrow Q \cup \{ (c, \delta_{c}, 1) \}$
                    \EndFor
                \Else
                    \State $[L, R] \Leftarrow$ children of $p$
                    \State $Q \Leftarrow Q \cup \{ (L, \delta^{+}_{L}, |L| - 1), (L_c, \delta_{L}, 1) \}$
                    \State $Q \Leftarrow Q \cup \{ (R, \delta^{+}_{R}, |R| - 1), (R_c, \delta_{R}, 1) \}$
                \EndIf
            \EndFor
        \EndLoop

        \State \textbf{return} $Q$
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm} 
    \caption{Depth-First Sieve($root$, $q$, $k$)} 
    \label{alg:methods:depth-first-sieve} 
    \begin{algorithmic}[0]
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find

        \State{$Q \Leftarrow [root]$, a min-priority queue by $\delta^{-}$}
        \State{$H \Leftarrow []$, a max-priority queue by $\delta$}

        \While{$|H| < k$ \textbf{or} $H.peek.\delta \geq Q.peek.\delta^{-}$}
            \While{$\neg (Q.peek$ is a leaf)}
                \State{$C \Leftarrow Q.pop$, the closest cluster}
                \State{$[L, R] \Leftarrow$ children of $C$}
                \State{$Q.push(L)$}
                \State{$Q.push(R)$}
            \EndWhile
            \State{$leaf \Leftarrow Q.pop$}
            \For{$p \in leaf$}
                \State{$H.push(p)$}
            \EndFor
            \While{$|H| > k$}
                \State{$H.pop$}
            \EndWhile
        \EndWhile
        \State Return $H$
    \end{algorithmic}
\end{algorithm}

@@

Due to their similarity, we combine the complexity analyses of the two Sieve methods into one section.
For these methods we, once again, use the terminology of tree-search and leaf-search.
Tree-search navigates the cluster tree and adds clusters to $Q$.
Leaf-search exhaustively searches some of the clusters in $Q$ to find the $k$ nearest neighbors.

@@

We perform some na\"{i}ve auto-tuning to select the optimal $k$-NN algorithm to use with a given dataset.
We start by taking the center of every cluster at a low depth, (e.g. 10), in the cluster tree as a query.
This gives us a small, representative sample of the dataset.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the three algorithms described in Section~\ref{sec:methods:knn-search}.
We select the fastest algorithm over all the queries as the optimal algorithm for that dataset and value of $k$.
Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.

@@

To elaborate on the augmentation process, we start with an original dataset from the ANN-benchmarks suite.
Let $X$ be the dataset, $d$ be its dimensionality, $\epsilon$ be a user-specified noise level, and $m$ be a user-specified integer multiplier.
For each datum $x \in X$, we create $m - 1$ new data points within a distance $\epsilon \cdot ||x||$ of $x$ where $||x||$ is the Euclidean distance from $x$ to the origin.
We construct a random vector $r$ of $d$ dimensions in the hyper-sphere of radius $\epsilon$ centered at the origin.
We then add $r$ to $x$ to get a new point $x'$.
Since $||r|| \leq \epsilon$, we have that $||x - x'|| \leq \epsilon$ (i.e., $x'$ is within a distance $\epsilon$ of $x$).
This produces a new dataset $X'$ with $|X'| = m \cdot |X|$.

### 2024-04-05 sections/3-datasets.tex
paper edits up until section 4.3

We benchmarked CAKES's algorithms against the state-of-the-art similarity search algorithms HNSW, ANNOY and FAISS-IVF, and two implementations of linear search.
In particular, we use FAISS-Flat, a Python implementation of linear search, to provide ground-truth values for recall of HNSW, ANNOY, and FAISS-IVF.
We use our own Rust implementation of linear search to provide ground-truth values for calculating recall of CAKES's algorithms.
We plot the throughput of these two implementations of linear search as a reference point for all other algorithms in Figures~\ref{fig:results:scaling-plots}

>>>>

We benchmarked CAKES's algorithms against the state-of-the-art similarity search algorithms HNSW, ANNOY and FAISS-IVF, and two implementations of linear search.
In particular, we use FAISS-Flat, a Python implementation of linear search, to provide ground-truth values for recall of HNSW, ANNOY, and FAISS-IVF.
We use our own Rust implementation of linear search to provide ground-truth values for calculating recall of CAKES's algorithms.
We plot the throughput of these two implementations of linear search as a reference point for all other algorithms in Figure~\ref{fig:results:scaling-plots}.
